The Ranked Nearest Neighbour Interchange space

### 2019-06-26 paper.tex
shorten proof: length of path if taxon deleted from all trees

A \amph{rooted binary phylogenetic tree} on a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \mapsto \mathcal{L}(T)$ a bijective mapping. \todo{do we need to define leaves and internal nodes, child, parent, subtree, cherry(defined below)?}
From now on, we assume $X = \{1,\ldots,n\}$ for a fixed number $n \in \mathbb N$ of leaves.
Within this paper we study \emph{ranked trees} that are determined by a pair consisting of a rooted binary phylogenetic tree $(t, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function.
A \emph{rank function} on $t$ maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
The rank of a node $v$ in a ranked tree will be denoted by $rank(v)$.
For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
An edge $e$ that has no leaf as child is an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if ${|rank(v) - rank(u)| = l}$.
Each internal edge defines a \emph{cluster} which is the set of taxa that are descending from the lower of both nodes bounding $e$. \todo{OK?}
In the literature, clusters often refer to sets of taxa descending from internal nodes in which case clusters as defined above coincide with non-trivial clusters.
Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
In particular, for a single taxon $x \in X$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
We write $mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

@@

The main purpose of this paper is understanding geometric properties of the $\rnni$ space.
Therefore, further notations need to be introduced.
If every internal node of a tree $T$ has at least one child that is a leaf, $T$ is called \emph{caterpillar tree}.
A \emph{path} $p$ between trees $T_0$ and $T_m$ in $\rnni$ is a sequence $T_0, T_1, \ldots, T_m$ of trees such that there is an edge $\{T_i, T_{i+1}\}$ for all $i=0,\ldots,m-1$.
For such a path $p$ we say the \emph{length} $|p|$ of $p$ is $m$.
A path $p$ between trees $T$ and $R$ is a \emph{shortest path} if for every path $r$ between $T$ and $R$, $|p| \leq |r|$.
The \emph{distance} $d(T,R)$ between trees $T$ and $R$ is the length of a shortest path between them.
If $p$ is a path only consisting of caterpillar trees we call it \emph{caterpillar path}.
For two caterpillar trees $T$ and $R$ a \emph{shortest caterpillar path} is a caterpillar path that is shortest among all caterpillar paths between $T$ and $R$.
The length of such a shortest caterpillar path is called \emph{caterpillar distance} and denoted by $d_c(T,R)$.
A set $\mathbb S$ of trees is called \emph{convex} if for all pairs of trees $T,R \in \mathbb S$, there exists a shortest path from $T$ to $R$ such that every tree on $p$ is element of $\mathbb S$.

@@

\begin{algorithm}[H]
\caption{FIND\_PATH($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
	\STATE $\hat{T} := T$
	\STATE $p := (T)$
	\FOR {$i = 1, \dots, n-2$}
		\STATE Let $C$ be the cluster associated with edge $(v,w)$ in $R$ with ${rank(v) = i < rank(w)}$
		\WHILE {$rank(mrca_{\hat{T}}(C))>i$}
			\STATE Update $\hat{T}$: Decrease $rank(mrca_{\hat{T}}(C))$ by an $\rnni$ move \label{alg:line:move_set_down}
			\STATE $p = p+\hat{T}$
		\ENDWHILE
	\ENDFOR
	\RETURN $p$
\end{algorithmic}
\end{algorithm}

@@

\begin{proof}
    Let us consider the maximum length of a path given by Algorithm~\ref{alg:find_path}.
    In worst case each of the $n-2$ clusters $C$ given by internal edges of $R$, represented by $i = 1, \dots, n-2$ in the algorithm, requires $n-1-i$ $\rnni$ moves for moving $mrca(C)$ down in $\hat{T}$.
	Hence, the maximum length of a path computed by that algorithm for any pair of trees is less or equal to ${\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}}$.
\end{proof}

Lemma~\ref{lemma:distance_delete_taxon} is central for some of the following induction proofs. \todo{Is it correct and if it is, do we need to prove the statement, or is it rather obvious?}

\todo{Change notations in this lemma+proof (less indices!!)}
\begin{lemma}
    Let $T_1$ and $T_2$ be two trees on $n+1$ taxa.
    Let $T_1|_n$ and $T_2|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T_1$ and $T_2$.
    Then it is $d(T_1|_n, T_2|_n) \leq d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ where $r_{T_i}(p_j)$ determines the rank of the parent of taxon $j$ in tree $T_i$.
    \label{lemma:distance_delete_taxon}
\end{lemma}

\begin{proof}
    Induction on $k = d(T_1,T_2)$:

    Basis($k = 1$):
    If rank of $p_{n+1}$ changes between $T_1$ and $T_2$, it is $d(T_1|_n, T_2|_n) = 0$ and $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = 1$ and therefore, $d(T_1|_n, T_2|_n) = 0 \leq 1 - 1 = d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
    If on the other hand the rank of $p_{n+1}$ does not change between $T_1$ and $T_2$, it is $d(T_1|_n, T_2|_n) = d(T_1,T_2) = 1$ and $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = 1$ and therefore, $d(T_1|_n, T_2|_n) = 1 \leq 1 - 0 = d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.

    Step($k \to k+1$):
    Let $T_1$ and $T_2$ be trees on $n+1$ taxa with distance $d(T_1,T_2) = k$.
    Let $T_2'$ be the last tree on a shortest path from $T_1$ to $T_2$, i.e. $d(T_1,T_2') = k$.
    We can apply the induction hypothesis on $T_1$ and $T_2'$ and see that $d(T_1|_n, T_2'|_n) \leq d(T_1, T_2') - |r_{T_1}(p_{n+1}) - r_{T_2'}(p_{n+1})| := k - m$.
    Now we differ two cases:
    \begin{enumerate}
        \item $p_{n+1}$ changes its rank between $T_2$ and $T_2'$ or is affected by the $\nni$ move is performed between $T_2$ and $T_2'$.

        Then the trees $T_2|_n$ and $T_2'|_n$ are equal. \todo{Is this obivous?}
        Therefore, it is $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m$.
        It is either $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m+1$, $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m$, or $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m-1$.
        In either case, we have $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m = (k+1) - (m+1) \leq d(T_1, T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.

        \item $p_{n+1}$ has same rank in $T_2$ and $T_2'$.

        In this case it is $d(T_1|_n,T_2|_n) \leq d(T_1|_n,T_2'|_n) + d(T_2'|_n, T_2|_n) \leq d(T_1|_n, T_2'|_n) + 1 \leq (k-m)+ 1$.
        It also follows that $|r_{T_1}(p_{n+1}) - r_{T_2'}(p_{n+1})| = |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ and therefore, we can conclude $d(T_1|_n,T_2|_n) \leq k-m+1 = k+1-m = d(T_1,T_2) -  |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
    \end{enumerate}
\end{proof}

@@

\begin{lemma}
    Let $T$ and $T'$ be two caterpillar trees.
    $T$ and $T'$ have distance $d(T,T') = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$ if, and only if, they have caterpillar distance $d_c(T,T') = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$. \todo{define $d_c$}
    \label{lemma:caterpillar_dist=diameter}
\end{lemma}

    % \todo{At the top of $T'$ it could also be $2,1$, or $t=1$}
    %
    % \begin{figure}[H]
    %     \centering
    %     \includegraphics[width=0.6\textwidth]{caterpillar_max_dist}
    %     \caption{Caterpillar trees $T$ and $T'$ with distance $d(T,T') = \Delta(\rnniu)$}
    %     \label{fig:caterpillar_max_distance}
    % \end{figure}

\begin{proof}
    Let $d_c(T,T')$ denote the length of a shortest caterpillar path from $T$ to $T'$, that is a caterpillar path that is shortest among all caterpillar paths from $T$ to $T'$.
    We need to prove that $d(T,T') = \Delta(\rnniu)$ if, and only if, $d_c(T,T') = \Delta(\rnniu)$.

    Let $T$ and $T'$ have distance $d(T,T') = \Delta(\rnniu)$.
    We can compute a  shortest caterpillar path from $T$ to $T'$ using Bubble Sort.
    In each loop $k=1, \ldots, n-2$ of the algorithm there can be at most $n-1-k$ exchanges of taxa.
    It follows that the maximum length of a path computed by Bubble Sort is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$ if it is $d(T,T') = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,T') = \Delta(\rnniu)$.

    Let now $T$ and $T'$ have caterpillar distance $d_c(T,T') = \Delta(\rnniu)$.
    We prove $d(T,T') = d_c(T,T')$ by induction on the number of taxa $n$ of $T$ and $T'$:
    We assume without loss of generality that $T$ is the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$.
    Otherwise, we re-label $T$ and $T'$.
    Let $T'$ be a caterpillar tree with distance $\Delta(\rnniu)$ to $T$.

@@

    Induction step: $n \to n+1$\\
    Induction hypothesis: For trees $R$, $R'$ on $n$ taxa it is $d_c(R,R') = d(R, R') = \frac{(n-1)(n-2)}{2} = \Delta(\rnniu_n)$.\\
    Let $T, T'$ be trees on $n+1$ taxa with caterpillar distance $\Delta(\rnniu_{n+1})$.
    When applying the Bubble Sort algorithm to compute a path between $T'$ and $T$, each pair of taxa that is compared in that algorithm exchanges, since there must be $\Delta(\rnniu)=\frac{(n-1)(n-2)}{2}$ moves between $T$ and $T'$.
    So we can conclude that taxon $n+1$ must be in the cherry of $T'$.
    Let us consider the trees $T|_n$ to $T'|_n$ that result from $T$ and $T'$ by deleting taxon $n+1$.
    When running Bubble Sort to compute a path from $T|_n$ to $T'|_n$ one can observe the same exchanges of taxa as on the Bubble Sort path from $T$ to $T'$, except for the $n-1$ ones that involve taxon $n+1$.
    Hence, the caterpillar distance is $d_c(T|_n, T'|_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
    So we can apply the induction hypothesis on $T|_n$ and $T'|_n$ and know that $d(T|_n,T'|_n) = \frac{(n-1)(n-2)}{2}$.

    With Lemma~\ref{lemma:distance_delete_taxon} we know that the distance between $T|_n$ and $T'|_n$ is at least $d(T|_n, T'|_n) \leq d(T,T') - |r_{T}(p_{n+1}) - r_{T'}(p_{n+1})|$.
    We also know $|r_{T}(p_{n+1}) - r_{T'}(p_{n+1})| = n-1$
    If we assume that there is a path from $T$ to $T'$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, it would follow that $d(T|_n, T'|_n) \leq d(T,T') - |r_{T}(p_{n+1}) - r_{T'}(p_{n+1})| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
    So it would be $d(T|_n, T'|_n) < \Delta(\rnniu_n)$.
    Since this is a contradiction to $d(T|_n, T'|_n) = \Delta(\rnniu_n)$, it must be $d(T,T') = \Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2} $.

@@

\begin{proof}
    We prove the Theorem by backwards induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T$ and $T'$.

    Induction basis: $l = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$\\
    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $T'$ with caterpillar distance $d_c(T,T') = \Delta(\rnniu)$, it is $d(T,T') = d_c(T,T')$, which means that there is a shortest path between $T$ and $T'$ that is a caterpillar path.

    Induction step: $l+1 \to l$\\
    Induction hypothesis: For each pair of caterpillar trees with caterpillar distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees, which means that the distance between these trees is $l+1$.
    Let $T$ and $T'$ be caterpillar trees with caterpillar distance $l$.
    Since $l < \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$, there is one loop $k$ in the Bubble Sort algorithm from $T'$ to $T$ where there are less than $n-2-k$ swaps of neighboured taxa.
    This means that there is a taxon $a$ that has parent with rank $n-2-k$ in $T$ and that does not need to move all the way up starting at the cherry in loop $k$.
    Therefore, there must be a taxon $b$ which has parent with rank one less than the parent of $a$ in the tree at the beginning of loop number $k$.
    If we assume that loop $k$ is the first loop where a single taxon is not moved by $n-k-2$ $\rnni$ moves, it follows that $a$ and $b$ must be neighboured taxa in $T'$.
    Let $T''$ be the caterpillar tree resulting from $T'$ by exchanging $a$ and $b$.
    It follows that $T'$ has caterpillar distance $d_c(T,T'') = d_c(T,T') + 1 = l+1$ to $T$ and $d(T',T'') = 1$.
    We can apply the induction hypothesis on $T$ and $T''$ and know that $d(T,T'') = l+1$.
    If it was $d(T,T') < l$, it would hold $d(T,T'') \leq d(T,T') + d(T',T'') < l + 1$ which contradicts the induction hypothesis.
    We can conclude that $d(T,T') = l$.

@@

\begin{proof}
	Let $T = (( \dots (n,1)2)\dots)n-1)$ and $R = (( \dots (n,n-1)n-2)\dots)1)$.
    According to Theorem~\ref{thm:caterpillar_convex}, there is a shortest path between $T$ and $R$ that is a caterpillar path.
    We can compute a shortest caterpillar path by using the modified version of Bubble Sort as explained above. \todo{Do we want to name this algorithm?}
    When running this algorithm on $T$ and $R$, taxon $i$ swaps $n-1-i$ times with its right neighbour.
    Therefore, the length of a shortest caterpillar path is ${\sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{(n-1)(n-2)}{2}}$.
\end{proof}

@@

% 	Let $T_1$ and $T_2$ be two caterpillar trees that both have a leaf labelled by $a$ as part of their cherry.
% 	There is a shortest paths from $T_1$ to $T_2$ that is a caterpillar path.

@@

% 	Let $T_1$ and $T_2$ be two caterpillar trees, both having a leaf labelled by $a$ as part of their cherry.
% 	Let $p$ be a shortest path from $T_1$ to $T_2$.
% 	We introduce a labelling of internal nodes of all trees on $p$ by first defining it for the caterpillar trees $T_1$ and $T_2$ and then explaining how the labelling is changed by $\rnni$ moves.

@@

% 	In $T_1$ and $T_2$ the internal node of the cherry is labelled by the label of its child that is not $a$.

@@

% 	Hence, there is always a shortest path between $T_1$ and $T_2$ that is a caterpillar path.

@@

\begin{theorem}
	There are trees $T_1,T_2$ in $\nni$ sharing a partition which is not shared by any intermediate tree on a shortest path from $T_1$ to $T_2$.
	\label{thm:split_nni}
\end{theorem}

>>>>

A \amph{rooted binary phylogenetic tree} on a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \mapsto \mathcal{L}(T)$ a bijective mapping. \todo{do we need to define leaves and internal nodes, child, parent, subtree, cherry(defined below)?}
From now on, we assume $X = \{1,\ldots,n\}$ for a fixed number $n \in \mathbb N$ of leaves.
Within this paper we study \emph{ranked trees} that are determined by a pair consisting of a rooted binary phylogenetic tree $(t, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function.
A \emph{rank function} on $t$ maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
The rank of a node $v$ in a ranked tree will be denoted by $rank(v)$ or $rank_T(v)$ if it is not obvious that we consider tree $T$.
For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
An edge $e$ that has no leaf as child is an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|rank(v) - rank(u)| = l$.
Each internal edge defines a \emph{cluster} which is the set of taxa that are descending from the lower of both nodes bounding $e$. \todo{OK?}
In the literature, clusters often refer to sets of taxa descending from internal nodes in which case clusters as defined above coincide with non-trivial clusters.
Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
In particular, for a single taxon $x \in X$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
We write $mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

@@

The main purpose of this paper is understanding geometric properties of the $\rnni$ space.
Therefore, further notations need to be introduced.
If every internal node of a tree $T$ has at least one child that is a leaf, $T$ is called \emph{caterpillar tree}.
A \emph{path} $p$ between trees $T_0$ and $T_m$ in $\rnni$ is a sequence $T_0, T, \ldots, T_m$ of trees such that there is an edge $\{T_i, T_{i+1}\}$ for all $i=0,\ldots,m-1$.
For such a path $p$ we say the \emph{length} $|p|$ of $p$ is $m$.
A path $p$ between trees $T$ and $R$ is a \emph{shortest path} if for every path $r$ between $T$ and $R$, $|p| \leq |r|$.
The \emph{distance} $d(T,R)$ between trees $T$ and $R$ is the length of a shortest path between them.
If $p$ is a path only consisting of caterpillar trees we call it \emph{caterpillar path}.
For two caterpillar trees $T$ and $R$ a \emph{shortest caterpillar path} is a caterpillar path that is shortest among all caterpillar paths between $T$ and $R$.
The length of such a shortest caterpillar path is called \emph{caterpillar distance} and denoted by $d_c(T,R)$.
A set $\mathbb S$ of trees is called \emph{convex} if for all pairs of trees $T,R \in \mathbb S$, there exists a shortest path from $T$ to $R$ such that every tree on $p$ is element of $\mathbb S$.

@@

\begin{algorithm}[H]
\caption{FIND\_PATH($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
	\STATE $\hat{T} := T$
	\STATE $p := (T)$
	\FOR {$i = 1, \dots, n-2$}
		\STATE Let $C$ be the cluster associated with edge $(v,w)$ in $R$ with $rank(v) = i < rank(w)$
		\WHILE {$rank(mrca_{\hat{T}}(C))>i$}
			\STATE Update $\hat{T}$: Decrease $rank(mrca_{\hat{T}}(C))$ by an $\rnni$ move \label{alg:line:move_set_down}
			\STATE $p = p+\hat{T}$
		\ENDWHILE
	\ENDFOR
	\RETURN $p$
\end{algorithmic}
\end{algorithm}

@@

\begin{proof}
    Let us consider the maximum length of a path given by Algorithm~\ref{alg:find_path}.
    In worst case each of the $n-2$ clusters $C$ given by internal edges of $R$, represented by $i = 1, \dots, n-2$ in the algorithm, requires $n-1-i$ $\rnni$ moves for moving $mrca(C)$ down in $\hat{T}$.
	Hence, the maximum length of a path computed by that algorithm for any pair of trees is less or equal to $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
\end{proof}

Lemma~\ref{lemma:distance_delete_taxon} is central for some of the following proofs using induction.

\begin{lemma}
    Let $T$ and $R$ be two trees on $n+1$ taxa.
    Let $T|_n$ and $R|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T$ and $R$.
    Then it is $d(T|_n, R|_n) \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1})|$.
    \label{lemma:distance_delete_taxon}
\end{lemma}

\begin{proof}
    On every path from $T$ to $R$ there are at least $|rank_T(parent(n+1)) - rank_R(parent(n+1})|$ $\rnni$ moves that change the rank of $parent(n+1)$ because each $\rnni$ move can change the rank of an internal node by at most one.
    In particular, this holds for a shortest path $p$ from $T$ to $R$.
    Furthermore, deleting taxon $n+1$ and its parent $p_{n+1}$ from all trees on $p$ results in a path $p|_n$ from $T|_n$ to $R|_n$.
    All moves that do not involve $p_{n+1}$ can be performed the same way and those that involved $p_{n+1}$ are just left out.
    We thus have a path $p|_n$ from $T|_n$ to $R|_n$ that has at least $|rank_T(parent(n+1)) - rank_R(parent(n+1})|$ $\rnni$ moves less than $p$, which results in $|p|_n| \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1})|$.
    \todo{Is this proof sufficient? Otherwise: induction}
    % Induction on $k = d(T,R)$:
    %
    % Basis($k = 1$):
    % If rank of $parent(n+1)$ changes between $T$ and $R$, it is $d(T|_n, R|_n) = 0$ and $|rank_T(parent(n+1)) - rank_R(parent(n+1))| = 1$ and therefore, $d(T|_n, R|_n) = 0 \leq 1 - 1 = d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
    % If on the other hand the rank of $parent(n+1}$ does not change between $T$ and $R$, it is $d(T|_n, R|_n) = d(T,R) = 1$ and $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = 1$ and therefore, $d(T|_n, R|_n) = 1 \leq 1 - 0 = d(T,R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    %
    % Step($k \to k+1$):
    % Let $T$ and $R$ be trees on $n+1$ taxa with distance $d(T,R) = k$.
    % Let $R'$ be the last tree on a shortest path from $T$ to $R$, i.e. $d(T,R') = k$.
    % We can apply the induction hypothesis on $T$ and $R'$ and see that $d(T|_n, R'|_n) \leq d(T, R') - |r_{T}(parent(n+1}) - r_{R'}(parent(n+1})| := k - m$.
    % Now we differ two cases:
    % \begin{enumerate}
    %     \item $parent(n+1}$ changes its rank between $R$ and $R'$ or is affected by the $\nni$ move is performed between $R$ and $R'$.
    %
    %     Then the trees $R|_n$ and $R'|_n$ are equal. \todo{Is this obivous?}
    %     Therefore, it is $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m$.
    %     It is either $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m+1$, $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m$, or $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m-1$.
    %     In either case, we have $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m = (k+1) - (m+1) \leq d(T, R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    %
    %     \item $parent(n+1}$ has same rank in $R$ and $R'$.
    %
    %     In this case it is $d(T|_n,R|_n) \leq d(T|_n,R'|_n) + d(R'|_n, R|_n) \leq d(T|_n, R'|_n) + 1 \leq (k-m)+ 1$.
    %     It also follows that $|r_{T}(parent(n+1}) - r_{R'}(parent(n+1})| = |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$ and therefore, we can conclude $d(T|_n,R|_n) \leq k-m+1 = k+1-m = d(T,R) -  |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    % \end{enumerate}
\end{proof}

@@

\begin{lemma}
    Let $T$ and $R$ be two caterpillar trees.
    $T$ and $R$ have distance $d(T,R) = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$ if, and only if, they have caterpillar distance $d_c(T,R) = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$. \todo{define $d_c$}
    \label{lemma:caterpillar_dist=diameter}
\end{lemma}

    % \todo{At the top of $R$ it could also be $2,1$, or $t=1$}
    %
    % \begin{figure}[H]
    %     \centering
    %     \includegraphics[width=0.6\textwidth]{caterpillar_max_dist}
    %     \caption{Caterpillar trees $T$ and $R$ with distance $d(T,R) = \Delta(\rnniu)$}
    %     \label{fig:caterpillar_max_distance}
    % \end{figure}

\begin{proof}
    Let $d_c(T,R)$ denote the length of a shortest caterpillar path from $T$ to $R$, that is a caterpillar path that is shortest among all caterpillar paths from $T$ to $R$.
    We need to prove that $d(T,R) = \Delta(\rnniu)$ if, and only if, $d_c(T,R) = \Delta(\rnniu)$.

    Let $T$ and $R$ have distance $d(T,R) = \Delta(\rnniu)$.
    We can compute a  shortest caterpillar path from $T$ to $R$ using Bubble Sort.
    In each loop $k=1, \ldots, n-2$ of the algorithm there can be at most $n-1-k$ exchanges of taxa.
    It follows that the maximum length of a path computed by Bubble Sort is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$ if it is $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \Delta(\rnniu)$.

    Let now $T$ and $R$ have caterpillar distance $d_c(T,R) = \Delta(\rnniu)$.
    We prove $d(T,R) = d_c(T,R)$ by induction on the number of taxa $n$ of $T$ and $R$:
    We assume without loss of generality that $T$ is the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$.
    Otherwise, we re-label $T$ and $R$.
    Let $R$ be a caterpillar tree with distance $\Delta(\rnniu)$ to $T$.

@@

    Induction step: $n \to n+1$\\
    Induction hypothesis: For trees $R$, $R'$ on $n$ taxa it is $d_c(R,R') = d(R, R') = \frac{(n-1)(n-2)}{2} = \Delta(\rnniu_n)$.\\
    Let $T, R$ be trees on $n+1$ taxa with caterpillar distance $\Delta(\rnniu_{n+1})$.
    When applying the Bubble Sort algorithm to compute a path between $R$ and $T$, each pair of taxa that is compared in that algorithm exchanges, since there must be $\Delta(\rnniu)=\frac{(n-1)(n-2)}{2}$ moves between $T$ and $R$.
    So we can conclude that taxon $n+1$ must be in the cherry of $R$.
    Let us consider the trees $T|_n$ to $R|_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
    When running Bubble Sort to compute a path from $T|_n$ to $R|_n$ one can observe the same exchanges of taxa as on the Bubble Sort path from $T$ to $R$, except for the $n-1$ ones that involve taxon $n+1$.
    Hence, the caterpillar distance is $d_c(T|_n, R|_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
    So we can apply the induction hypothesis on $T|_n$ and $R|_n$ and know that $d(T|_n,R|_n) = \frac{(n-1)(n-2)}{2}$.

    With Lemma~\ref{lemma:distance_delete_taxon} we know that the distance between $T|_n$ and $R|_n$ is at least $d(T|_n, R|_n) \leq d(T,R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    We also know $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = n-1$
    If we assume that there is a path from $T$ to $R$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, it would follow that $d(T|_n, R|_n) \leq d(T,R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
    So it would be $d(T|_n, R|_n) < \Delta(\rnniu_n)$.
    Since this is a contradiction to $d(T|_n, R|_n) = \Delta(\rnniu_n)$, it must be $d(T,R) = \Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2} $.

@@

\begin{proof}
    We prove the Theorem by backwards induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T$ and $R$.

    Induction basis: $l = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$\\
    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $R$ with caterpillar distance $d_c(T,R) = \Delta(\rnniu)$, it is $d(T,R) = d_c(T,R)$, which means that there is a shortest path between $T$ and $R$ that is a caterpillar path.

    Induction step: $l+1 \to l$\\
    Induction hypothesis: For each pair of caterpillar trees with caterpillar distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees, which means that the distance between these trees is $l+1$.
    Let $T$ and $R$ be caterpillar trees with caterpillar distance $l$.
    Since $l < \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$, there is one loop $k$ in the Bubble Sort algorithm from $R$ to $T$ where there are less than $n-2-k$ swaps of neighboured taxa.
    This means that there is a taxon $a$ that has parent with rank $n-2-k$ in $T$ and that does not need to move all the way up starting at the cherry in loop $k$.
    Therefore, there must be a taxon $b$ which has parent with rank one less than the parent of $a$ in the tree at the beginning of loop number $k$.
    If we assume that loop $k$ is the first loop where a single taxon is not moved by $n-k-2$ $\rnni$ moves, it follows that $a$ and $b$ must be neighboured taxa in $R$.
    Let $R'$ be the caterpillar tree resulting from $R$ by exchanging $a$ and $b$.
    It follows that $R$ has caterpillar distance $d_c(T,R') = d_c(T,R) + 1 = l+1$ to $T$ and $d(R,R') = 1$.
    We can apply the induction hypothesis on $T$ and $R'$ and know that $d(T,R') = l+1$.
    If it was $d(T,R) < l$, it would hold $d(T,R') \leq d(T,R) + d(R,R') < l + 1$ which contradicts the induction hypothesis.
    We can conclude that $d(T,R) = l$.

@@

\begin{proof}
	Let $T = (( \dots (n,1)2)\dots)n-1)$ and $R = (( \dots (n,n-1)n-2)\dots)1)$.
    According to Theorem~\ref{thm:caterpillar_convex}, there is a shortest path between $T$ and $R$ that is a caterpillar path.
    We can compute a shortest caterpillar path by using the modified version of Bubble Sort as explained above. \todo{Do we want to name this algorithm?}
    When running this algorithm on $T$ and $R$, taxon $i$ swaps $n-1-i$ times with its right neighbour.
    Therefore, the length of a shortest caterpillar path is \sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{(n-1)(n-2)}{2}.
\end{proof}

@@

% 	Let $T$ and $R$ be two caterpillar trees that both have a leaf labelled by $a$ as part of their cherry.
% 	There is a shortest paths from $T$ to $R$ that is a caterpillar path.

@@

% 	Let $T$ and $R$ be two caterpillar trees, both having a leaf labelled by $a$ as part of their cherry.
% 	Let $p$ be a shortest path from $T$ to $R$.
% 	We introduce a labelling of internal nodes of all trees on $p$ by first defining it for the caterpillar trees $T$ and $R$ and then explaining how the labelling is changed by $\rnni$ moves.

@@

% 	In $T$ and $R$ the internal node of the cherry is labelled by the label of its child that is not $a$.

@@

% 	Hence, there is always a shortest path between $T$ and $R$ that is a caterpillar path.

@@

\begin{theorem}
	There are trees $T,R$ in $\nni$ sharing a partition which is not shared by any intermediate tree on a shortest path from $T$ to $R$.
	\label{thm:split_nni}
\end{theorem}
